---
title: "A Gentle Introduction to R"
subtitle: 'EXTRAS'
author: "Jonathan Whiteley"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    theme: "Boadilla"
header-includes: |
  \setbeamertemplate{itemize item}[circle]
  \newcommand{\Rlogo}{\includegraphics[height=1em]{`r file.path(R.home("doc"), "html", "logo.jpg")`}}
  \newcommand{\R}{\texttt{R}}
  \newcommand{\highlight}[1]{\StringTok{#1}}
  \newcommand{\important}[1]{\AlertTok{#1}}
  \newcommand{\fade}[1]{\textcolor[rgb]{0.66,0.66,0.66}{#1}}
  \newcommand{\annote}[1]{{\footnotesize #1}}
  \newcommand{\obj}[1]{\textit{texttt{\VariableTok{#1}}}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Matrix math

- $\R$ can do _matrix math_ --- which are uses in many statistical procedures
  - But the _syntax_ is different from the usual math operators
- Using a regular multiplication symbol (`*`) results in *element*-wise multiplication
  - each _element_ (item) in matrix1 is multiplied by the corresponding _element_ in matrix2, etc.
```{r results='hide'}
c(1, 2, 3) * c(3, 2, 1)
```
- _Matrix multiplication_ is specified by this operator: `%*%`
```{r results='hide'}
c(1, 2, 3) %*% c(3, 2, 1)
```


# Symbolic _Variables_

- You can store values (*objects*) in symbolic variables (*names*)\
  using an *assignment operator*
<!--
||| empty header row, because most Markdown parsers require a header row 
|:---|-|
|`->`|assign the *value* on the **left** to the *name* on the **right**|
|`<-`|assign the *value* on the **right** to the *name* on the **left**|
|`=` |assign the *value* on the **right** to the *name* on the **left**|
-->
```{=latex}
% Latex table, because I want to use LaTeX commands to format text,
% which somehow causes extra space in the tex output, if used in a Markdown table.
\begin{longtable}[c]{@{}ll@{}}
\toprule()
\endhead
\OperatorTok{\texttt{-\textgreater{}}} & assign the \emph{value} on the \textbf{left} to
the \emph{name} on the \textbf{right} \\
\OperatorTok{\texttt{\textless{}-}} & assign the \emph{value} on the \textbf{right} to
the \emph{name} on the \textbf{left} \\
\OperatorTok{\texttt{=}} & assign the \emph{value} on the \textbf{right} to
the \emph{name} on the \textbf{left} \\
\bottomrule()
\end{longtable}
```

- '`<-`' is preferred, because it is unambiguous\
  (to people _and_ to $\R$)
- '`=`' is not allowed in certain situations\
  (e.g., when [surrounded by other expressions](https://stackoverflow.com/a/51564252))
  - '`=`' is also used to set *argument* *values* in *function calls*,\
  which is a different meaning and its most common use.

- You can also use the `assign` *function* (advanced):

```{r, results='hide'}
assign('x', 3)    # assign the value 3 to the variable ‘x’
```

<!--
https://stackoverflow.com/questions/1741820/what-are-the-differences-between-and-assignment-operators
-->

# Variable / Object Names

- In $\R$, all variables are *objects*
  - In $\R$, **everything** is an *object*
  
- Object names can include:\
  (depending on the language or *locale*)

|||<!-- empty header row, because most Markdown parsers require a header row -->
|:----------|---------|
|letters    |`a-z A-Z`|
|numbers    |`0-9`    |
|periods    |`.`      |
|underscores|`_`      |

- Names _should begin with a letter_

```{r, results='hide'}
A <- 10
B = 10 * 10
log(A) -> A_log
B.seq <- 1:B
assign('x', 3)
```


# Object Names: Details

Names can start with a **letter** or a **period** _(more on this later)_

```{r, results='hide'}
myvar  <- T
.myvar <- T
```
  
but anything else triggers an \important{error}
```{r, eval=FALSE}
0myvar <- F
_myvar <- F
my var <- F
```

## <!-- empty subheader to separate the next part -->

For more information about object names in \R{}, see:

- [Section 1.8](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#R-commands_003b-case-sensitivity-etc) of 
  ['An Introduction to R'](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)
  
- [Section 2.1.3](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Language-objects) of
  ['The R Language Definition'](https://cran.r-project.org/doc/manuals/r-release/R-lang.html)



# Object Names: Hidden

- Names starting with a period (`.`) are special and normally hidden from users.

```{r, results='hide'}
ls()
ls(all.names = TRUE)
```

  - Names starting with a period are used by packages or the system for special objects that users should not interact with directly.
  - Such objects may not behave as expected with common commands, such as `ls()` (above).
  - Therefore, most users should avoid doing this unless they know what they are doing and have a good reason to do so.


# Object Names: Advanced

- 'Valid' names following the rules above can be referred to easily in code.

- Names with any character are actually possible, but must be quoted with backticks (`` ` ``)
  - **This is not recommended practice**, but occasionally useful when you need to refer to an element of an object, such as list items or data frame columns, that have non-standard names.

```{r}
`(my) [strange] {variable} 'name' "!@#$"` <- T
print(`(my) [strange] {variable} 'name' "!@#$"`)
```


