---
title: "A Gentle Introduction to R"
subtitle: 'EXTRAS'
author: "Jonathan Whiteley"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    theme: "Boadilla"
header-includes: |
  \newcommand{\Rlogo}{\includegraphics[height=1em]{`r file.path(R.home("doc"), "html", "logo.jpg")`}}
  \newcommand{\R}{\texttt{R}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Matrix math

- $\R$ can do _matrix math_ --- which are uses in many statistical procedures
  - But the _syntax_ is different from the usual math operators
- Using a regular multiplication symbol (`*`) results in *element*-wise multiplication
  - each _element_ (item) in matrix1 is multiplied by the corresponding _element_ in matrix2, etc.
```{r results='hide'}
c(1, 2, 3) * c(3, 2, 1)
```
- _Matrix multiplication_ is specified by this operator: `%*%`
```{r results='hide'}
c(1, 2, 3) %*% c(3, 2, 1)
```


# Variable Names: Details

- Variable names containing only the following are easiest to work with:

|||<!-- empty header row, because most Markdown parsers require a header row -->
|:----------|---------|
|letters    |`a-z A-Z`|
|numbers    |`0-9`    |
|periods    |`.`      |
|underscores|`_`      |

- Names can start with a **letter** or a **period** _(more on this later)_, but anything else triggers an error

```{r, results='hide'}
myvar = T
.myvar = T
```
  
These will fail:
```{r, eval=FALSE}
0myvar = F
_myvar = F
my var = F
```


# Variable Names: Advanced

- 'Valid' names following the rules above can be referred to easily in code.

- Names with any character are actually possible, but must be quoted with backticks (```)
  - **This is not recommended practice**, but occasionally useful when you need to refer to an element of an object, such as lists and data frames, that have non-standard names.

```{r}
`(my) [strange] {variable} 'name' "!@#$"` = T
print(`(my) [strange] {variable} 'name' "!@#$"`)
```


# Variable Names: Hidden

- Variable names starting with a period (`.`) are special and normally hidden from users.

```{r, results='hide'}
ls()
ls(all.names = TRUE)
```

  - Such variables are used by packages or the system for special values that users should not interact with directly.
  - Such variables may not behave as expected with common commands, such as `ls()` (above).
  - Therefore, most users should avoid doing this unless they know what they are doing and have a good reason to do so.

