---
title: "A Gentle Introduction to R"
author: "Jonathan Whiteley"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    theme: "Boadilla"
header-includes: |
  \newcommand{\Rlogo}{\includegraphics[height=1em]{`r file.path(R.home("doc"), "html", "logo.jpg")`}}
  \newcommand{\R}{\texttt{R}}
  \setbeamertemplate{itemize item}[circle]

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pop Quiz

```{=latex}
{\footnotesize We will review these \textit{at the end}, so you can see how much you have learned.}
```

+ What does 'CRAN' stand for?
+ Why is it called '$\R$'?
+ How can you use $\R$ _interactively_?
+ How do you find out what a function does & how to use it?
+ How do you store values to re-use later?
+ True or False: Warnings can be ignored, but an Error means I made a mistake.
+ True or False: Error messages will tell me how to fix the problem.


# Learning Objectives

* Get familiar with the ![]( `r file.path(R.home("doc"), "html", "logo.jpg")`){height=1em} _interface_
* Enter _commands_
  * input & output: using $\R$ interactively
  * use some common _functions_
* Understand _Errors_ & _Warnings_
* Use technical _terms_ for $\R$ concepts
* How to get Help

# Why is it called $\Rlogo$?

- **$\R$** started as an _open-source_ implementation of the  
  **`S`** statistical computing language (S-PLUS)
  - `S` was created at Bell Laboratories in 1976
  - $\R$ was based on the `S` syntax (mostly v3), but works very differently "under the hood".
- **$\R$** was created by **R**oss Ihaka and **R**obert Gentleman at the University of Aukland in the early 1990s.

# The $\Rlogo$ Inteface

- $\R$ has a slightly different interface for each **O**perating **S**ystem (OS)
- GUI = **G**raphical **U**ser **I**nterface
- In every case, you interact with R primarily using a _command line_
  - aka “Question-and-Answer Model”
  - You ask $\R$ to do something (a _command_),  
    \qquad{} and $\R$ tells you the answer (_result_).
  - Instructions are given to $\R$ using the _R language_.



# Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

```{r}
1 + 2
```

# Slide with Plot

```{r pressure}
plot(pressure)
```

