---
title: "Final Activity: read, clean & combine files, then analyze"
subtitle: "A Short Introduction to Working With Data in R"
output: html_document # change to 'html_notebook' to automatically save html with output
---

Remember:

-   *code in an R Markdown document uses the file location as the working directory.*

-   *Output appears in this document, not the console.*

## Libraries

```{r libraries}
library(tidyverse)
```

# Load

Load *all* the data files in this directory: `../data/activity` (".." in the path refers to the *parent directory* of where this file is)

```{r load data}
list.files( file.path("..", "data", "activity") )
```

Data are from the "World Health Organization Global Tuberculosis Report", with counts of new cases in each year for a subset of countries, by group (method of diagnosis, gender, and age)\
(see `?tidyr::who` for details).

# Clean

Check & clean the files, as appropriate.

```{r}

```

## Process

Combine the data into a single data frame with:

-   The total number of new TB cases in each year and country
-   The country's population in that year
-   The rate of new cases (total cases / population)

```{r}

```

# Analyze

Can you answer these questions with the data?

1.  Which country had the highest all-time rate of new cases?
2.  In which year was it?
3.  Which country has had the *lowest* average rate since 2006 (including that year)?
4.  Which country had the highest & lowest *growth rate* in *cases* since 2006 (including that year)?
5.  What changed in 2006?

```{r}

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (*Cmd+Option+I* on mac*)*.

# Save

*optional*: Save the data in the parent directory (`data`)

```{r}
out_path <- file.path("..", "data")
```
