---
title: "A Short Introduction to Working With Data in R"
author: "Jonathan Whiteley"
date: "`r Sys.Date()`"
output: 
  beamer_presentation: 
    latex_engine: xelatex # ensures certain characters (like '^') are rendered correctly
    theme: Boadilla
    colortheme: default
    highlight: ../shared/my_tango.theme
    slide_level: 2
    toc: false
    keep_tex: yes
    includes:
      in_header: "../shared/preamble-include.tex"
fontsize: 11pt
urlcolor: darkblue   # defined in LaTeX preamble (in_header includes, above)
linkcolor: darkblue  # defined in LaTeX preamble (in_header includes, above)
params:
  is_virtual: TRUE    # is this presentation for a virtual (T) or in-person (F) environment?
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 60)  # to fit output on slides
```

## Prerequisites

-   Access to a copy of the [![](`r file.path(R.home("doc"), "html", "logo.jpg")`){height=1em}](https://www.r-project.org) software

    -   i.e., a "binary executable"

    -   Go to [*www.r-project.org*](https://www.r-project.org) to get a copy, \
        or ask your system administrator.

- Tidyverse packages installed on the same system as $\R$

  - Please run this command in $\R$ *before* the workshop:
    ```{r install_tidyverse, eval=FALSE}
    install.packages("tidyverse")
    ```

-   Knowledge of common mathematical operations: arithmetic, logarithms, etc.

-   Knowledge of basic $\R$ concepts, such as *variables*, *objects*, *operators*, *functions*, *packages*, etc.

    -   This is covered in the first workshop: "A Gentle Introduction to R"

<!-- R logo: ![](`r file.path(R.home("doc"), "html", "logo.jpg")`){height=1em} -->


## Learning Objectives

- Load tabular data into R
- Explore data to check that it was loaded correctly
- Export data from R to external files
- Data frames
- Clean data
  - Add & change columns
  - Edit values systematically
  - Change data types
- Tidy data
  - Change the _shape_ of a data frame
- Re-use code, reproducible results, automated reports
  - Scripts
  - R Markdown, R Notebooks

# Welcome

## Pop Quiz

\annote{\fade{We will review these \textit{at the end}, so you can see how much you have learned.}}

+ 

```{r welcome, echo=FALSE, results='asis'}
if (params$is_virtual) {
  cat("\\ 
  \n### Answer in the chat:
  \n
  \nWhat is your favourite emoji?  Why do you like to use it so much?
  ")
}
```


## Introductions

- Name
- Pronouns
- Job title, role
- _optional_: a favourite childhood treat or candy?

<!-- -->

- What are you hoping to learn most in today's workshop?


# Loading data into $\R$


# Exploring your data

## Data frames

# Saving data outside $\R$


# Re-using your code: scripts and other files


# The `tidyverse` collection of packages


# Clean data


# Tidy data


# Review

## Exercise


## [Quiz](#pop-quiz) Review

<!-- copy questions from [Pop Quiz][], above. -->


::: notes
`## Answers (for discussion)`

:::




# Backmatter

## References

::: notes
<https://cran.rapporter.net/doc/contrib/de_Jonge+van_der_Loo-Introduction_to_data_cleaning_with_R.pdf>
:::
