---
title: "A Gentle Introduction to R"
author: "Jonathan Whiteley"
date: "`r Sys.Date()`"
output: ioslides_presentation
  #beamer_presentation:
    #theme: "Boadilla"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pop Quiz

+ What does 'CRAN' stand for?
+ Why is it called 'R'?
+ How can you use R _interactively_?
+ How do you find out what a function does & how to use it?
+ How do you store values to re-use later?
+ True or False: Warnings can be ignored, but an Error means I made a mistake.
+ True or False: Error messages will tell me how to fix the problem.


## Learning Objectives

* Get familiar with the R _interface_
* Enter _commands_
  * input & output: using R interactively
  * use some common _functions_
* Understand _Errors_ & _Warnings_
* Use technical _terms_ for R concepts
* How to get Help

## Why is it called 'R'?

- **`R`** started as an _open-source_ implementation of the 
  **`S`** statistical computing language (S-PLUS)
  - `S` was created at Bell Laboratories in 1976
  - `R` was based on the `S` syntax (mostly v3), but works very differently "under the hood".
- **`R`** was created by **R**oss Ihaka and **R**obert Gentleman at the University of Aukland in the early 1990s.

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

```{r}
1 + 2
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

